---
title: "Bike_data_analysis"
author: "Brian Leslie"
date: '2022-11-08'
output: pdf_document
---

### We start by importing the packages we need

```{r}
library(tidyverse)  # Helps wrangle data
library(lubridate)  # Helps wrangle date attributes
library(ggplot2)  # Helps visualize data
library(dplyr)   # Helps manipulate data
library(readr)    # Helps read csv files
```

### Preparing Our Data

```{r}
setwd("c:/Users/bsoxl/OneDrive/Documents")
X202110 <- read_csv("bike_data/202110-divvy-tripdata.csv")
X202111 <- read_csv("bike_data/202111-divvy-tripdata.csv")
X202112 <- read_csv("bike_data/202112-divvy-tripdata.csv")
X202201 <- read_csv("bike_data/202201-divvy-tripdata.csv")
X202202 <- read_csv("bike_data/202202-divvy-tripdata.csv")
X202203 <- read_csv("bike_data/202203-divvy-tripdata.csv")
X202204 <- read_csv("bike_data/202204-divvy-tripdata.csv")
X202205 <- read_csv("bike_data/202205-divvy-tripdata.csv")
X202206 <- read_csv("bike_data/202206-divvy-tripdata.csv")
X202207 <- read_csv("bike_data/202207-divvy-tripdata.csv")
X202208 <- read_csv("bike_data/202208-divvy-tripdata.csv")
X202209 <- read_csv("bike_data/202209-divvy-publictripdata.csv")

year_data <- rbind(X202110,X202111,X202112,X202201,X202202,X202203,
 X202204,X202205,X202206,X202207,X202208,X202209)
View(year_data) # Combined data tables together
colnames(year_data) # List of column names
nrow(year_data)  # Shows number of rows
dim(year_data)   # Dimensions of the data frame
str(year_data)   # Shows data types of columns
summary(year_data) # Statistical suummary of data
year_data <- year_data %>%
  mutate(member_casual = recode(member_casual
                                ,"Subscriber"="member"
                                ,"Customer"="casual"))

```

### Cleaning and Processing Our Data

##### The data appears to have already been cleaned regarding the member

##### status column since no rows were removed.

```{r}
table(year_data$member_casual) #only two variables
```

##### Add columns that list the date, month, day, and year of each ride

```{r}
year_data$date <- as.Date(year_data$started_at)
year_data$month <- format(as.Date(year_data$date), "%m")
year_data$day <- format(as.Date(year_data$date), "%d")
year_data$year <- format(as.Date(year_data$date), "%Y")
year_data$day_of_week <-format(as.Date(year_data$date), "%A")
```

##### We will now add a ride_length column in seconds

```{r}
year_data$ride_length <- difftime(year_data$ended_at,year_data$started_at)
summary(year_data$ride_length) #we need to convert from factor to numeric
year_data$ride_length <- as.numeric(as.character(year_data$ride_length))
is.numeric(year_data$ride_length)
summary(year_data$ride_length) #we see a negative minimum, we need to clean
year_data_v2 <- year_data[!(year_data$start_station_name =="HQ QR"|year_data$ride_length<0),]
summary(year_data_v2$ride_length) # now we can describe the data will basic statistical analysis
```

### Analyzing Our Data

```{r}
aggregate(year_data_v2$ride_length ~ year_data_v2$member_casual, FUN = mean)
aggregate(year_data_v2$ride_length ~ year_data_v2$member_casual, FUN = median)
aggregate(year_data_v2$ride_length ~ year_data_v2$member_casual, FUN = max)
aggregate(year_data_v2$ride_length ~ year_data_v2$member_casual, FUN = min)
```

##### See the average ride time by each day for members vs casual users

```{r}
aggregate(year_data_v2$ride_length ~ year_data_v2$member_casual + year_data_v2$day_of_week, FUN = mean)
```

#####Now lets order the days

```{r}
year_data_v2$day_of_week <- ordered(year_data_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(year_data_v2$ride_length ~ year_data_v2$member_casual + year_data_v2$day_of_week, FUN = mean)
```

##### Now lets find the average ride time by day by member status

```{r}
year_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% # Creates weekday field
  group_by(member_casual, weekday) %>% # Groups by user type and day
  summarise(number_of_rides=n() # Calculations
            ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) # Sorting
```

##### Now we visualize number of rides by user type

```{r}
year_data_v2 %>%
  mutate(weekday =wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides=n()
            ,average_duration=mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday,y=number_of_rides, fill=member_casual))+
  geom_col(position="dodge")
```

##### Now lets compare trip duration by user type

```{r}
year_data_v2 %>%
  mutate(weekday =wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides=n()
            ,average_duration=mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday,y=average_duration, fill=member_casual))+
  geom_col(position="dodge")
```

##### Saving our new data frame

```{r}
write.csv(year_data_v2,"bike_data/.csv")
```
